### 2019-10-14 ch1_introduction/content.tex
update

% KDP: je zegt niets over epigenetica, die min of meer bepaald welke gebieden in het DNA worden vertaald (dit leidt je misschien te ver...)
During the process of \textbf{transcription} that takes place in the cell nucleus, a complementary RNA copy is transcribed from the template DNA. The initial RNA transcript is a precursor messenger RNA (pre-mRNA) that needs to undergo series of maturation steps to ultimately form the mature messenger RNA (mRNA). This maturation includes pre-mRNA \textbf{splicing} to remove non-protein coding intervening sequences (the introns) and to join the neighbouring protein-coding sequences (the exons). A single pre-mRNA can be alternatively spliced to generate multiple forms of mRNAs that will result in the production of multiple protein isoforms. This process of alternative splicing is essential to generate more than 100'000 different proteins starting from just 20'000 genes\cite{nilsen_expansioneukaryoticproteome_2010}.

@@

\subsection{Cell types}
\textit{Homo sapiens} like to categorise everything they encounter, and so too have they conceptualised groups of cells called "cell types". The human body contains more than 200 different cell types that are classified into four groups: epithelial, connective, muscle, and nervous. This however, is a major underestimation of the real number of cell types. Neurons, for instance, that are known to be extremely diverse, are estimated to reach numbers above 10,000 different types\cite{strachan_humanmoleculargenetics_2011}.

@@

Flow cytometry\cite{fulwyler_electronicseparationbiological_1965} circumvents imaging and segmentation issues by running a steady stream of cells through a laser and measuring the amount of fluorescent light scattered from those cells. 
Since cells need to be suspended in a buffer, flow cytometry is particularly useful for analysing non-adherent cells such as the many different immune cells in blood. However, many protocols already exist to extract viable single cells from tissues and tumours\cite{leelatian_preparingviablesingle_2017}.
Conventional flow cytometry devices enable to measure protein expression levels of millions of cells using up to eight different antibody fluorochromes simultaneously, while state-of-the-art instrumentation allows detection of up to 27 biomarkers simultaneously\cite{cossarizza_guidelinesuseflow_2017}.

@@

\subsubsection{Omics}
On the other side of the spectrum are the so-called "omics" technologies. "Omics"\footnote{The etymology of "omics" is quite interesting\cite{yadav_wholenesssuffixomics_2007}.} is a collective term for profiling all molecules of a particular type in a high-throughput manner.
There are many types of "omics", such as genomics, transcriptomics and proteomics. Genomics studies the complete DNA sequence of an organism's genome, while transcriptomics and proteomics study the RNA transcripts and proteins, respectively.

@@

The rapidly advancing field of single-cell omics harbours exceptional opportunities to discover new aspects of biology and redefine existing knowledge. Some of these opportunities lie in efforts like the HCA consortium. They have set out to redefine all human cell types in both terms of their gene expression and location, and as well as the developmental trajectories connecting the different cell types. As part of this endeavour, the consortium will likely profile the whole transcriptomes of tens or even hundreds of millions of cells.

@@

\subsection{Trajectory inference}
While clustering methods divide cells into distinct groups, trajectory inference (TI) methods acknowledge that cells are dynamic entities which transition from once cellular state to another via various dynamic processes. Rather than making distinct groups, TI methods allow studying dynamic processes by reconstructing the topology of a dynamic process as a trajectory, and map the cells onto that trajectory (Figure~\ref{fig:comp_tools_ti}A). Similar to clustering, by colouring the cells according to the expression of genes known to be involved in the dynamic process of interest, the milestones in the trajectory can be annotated (Figure~\ref{fig:comp_tools_cluster}B).

\begin{figure}[htb!]
	\centering
	\includegraphics[width=.6\linewidth]{fig/comptools2/comptool_3_trajectoryinference.pdf} 
	\caption{
		\textbf{Trajectory inference for single-cell omics data.}
		\textbf{A:} During a dynamic process cells pass through several transitional states, characterized by different waves of transcriptional, morphological, epigenomic and/or surface marker changes\cite{enver_stemcellstates_2009}. TI methods provide a straightforward and unbiased approach to identifying and correctly ordering different transitional stages. 
		\textbf{B:} By overlaying gene expression levels on a dimensionality reduction, the milestones can be annotated to allow better interpretation of the cellular heterogeneity.
	}
	\label{fig:comp_tools_ti}
\end{figure}

TI methods typically first require some form of dimensionality reduction prior to inferring a trajectory from the dataset (Figure~\ref{fig:ti_metro}). Subsequently, a trajectory modelling step then operates in this reduced space, aiming to identify cell states, constructing a trajectory through the different states, and projecting the cells back on to the trajectory.

A common way to classify TI methods is by the types of trajectories they can infer\cite{saelens_comparisonsinglecelltrajectory_2019}. About half of TI methods specialise in inferring linear or cyclic trajectories (i.e. they order the cells). Others model the trajectory as a rooted tree, allowing for one or more bifurcations to occur. Only a few methods are able to infer more generalised trajectories containing disconnected subgraphs or cycles.

\begin{figure}[htb!]
	\centering
	\includegraphics[width=.9\linewidth]{fig/ti_metro} 
	\caption{
		\textbf{TI methods use several common building blocks and can be organized in a unifying modular framework.} Overall, TI consists of two steps. In the first step dimensionality reduction techniques such as manifold learning, clustering, or graph-based methods are used to convert the dataset to a more simplified representation. This representation of the data then allows the trajectory itself to be more easily modelled in a second step. In this step, the trajectory itself is found within the data using both graph-based and curve-based approaches, after which the cells themselves can be ordered using a variety of methods.
	}
	\label{fig:ti_metro}
\end{figure}

@@

Network inference (NI) methods predict which genes are regulated by which transcription factors (Figure~\ref{fig:comp_tools_ni}). Before single-cell omics, these methods rely on multiple experiments, amongst which perturbation and time-series experiments, to predict the effect each transcription factor has on the up- or downregulation of a gene.

One of the main advantages of single-cell omics is the heterogeneity between cells caused by naturally occurring biological randomness\cite{padovan-merhar_usingvariabilitygene_2013} can be exploited to infer regulatory interactions between TFs and their target genes at much lower costs.

@@

\section{Research context and objectives}
Recent technological advancements in profiling single cells are having significant repercussions in many fields of biology. Profiling thousands of individual cells in a genome-wide manner provides opportunities to study cell heterogeneity and dynamics, for example inferring mechanisms for cellular development or intercellular communication. 
Hundreds of new software tools were developed\cite{zappia_exploringsinglecellrnaseq_2018} to perform these new types of analyses, or to fit existing analytical tools to deal with new data characteristics (e.g. differential expression, dimensionality reduction, normalisation).

One major shortcoming during the advent of single-cell omics was that majority of the newly developed computational tools were not quantitatively and comparatively evaluated. Rather, they relied on anecdotal evidence to demonstrate its usefulness. This issue is not the result of the tool developer's malevolence, but instead of the lack of data required to perform such comprehensive benchmarks.
%TODO: also mention guidelines for users.

@@

In this work, we aim to speed up scientific progress in single-cell omics by providing tools both for end-users and developers alike. For developers of computational approaches, we provide tools and guidelines for benchmarking their method on real and synthetic data. For end-users we develop new tools and guidelines for analysing dynamic processes by inferring trajectories and gene regulatory networks. These contributions are discussed in the following chapters:

\begin{itemize}
	\item We develop benchmarking strategies for assessing the performance of computational tools constrained by low availability of novel types of real single-cell data (Chapter~\ref{chap:dyngen}). \textit{In silico} simulations of individual cells are used to help kick-start emerging domains much more safely and allow anticipation of future technological developments by already developing computational tools.
	\item We apply this strategy to perform a comparison of TI methods (Chapter~\ref{chap:dynbenchmark}). Trajectory inference is one of the largest categories of all the novel single-cell omics tools, yet a comprehensive and quantitative study of the advantages and disadvantages of the numerous tools was hitherto lacking. We provide a set of guidelines for end-users wishing to infer trajectories. We also make our pipeline, datasets, metrics, and containerised wrappers of TI methods publicly available for developers to use.
	\item We developed dyno, a toolkit to easily infer, visualise and interpret single-cell trajectories using more than 50 different TI methods (Chapter~\ref{chap:dyno}). dyno provides downstream analysis such as: visualising a trajectory in a low-dimensional space or a heatmap, detecting genes differentially expressed at different stages of the tragectory, comparing multiple trajectories in a common dimensionality reduction, and manipulating the trajectory (e.g. adding directionality or adding annotation).
	\item We introduce a novel TI method specialised in inferring linear trajectories (Chapter~\ref{chap:scorpius}). Despite linear TI being the most simple but commonly used form of trajectory inference, the benchmark demonstrated that most TI methods are not capable of producing accurate models of linear datasets.
	\item We invent a new type of NI method capable of inferring the GRN of individual cells (Chapter~\ref{chap:bred}). We demonstrate this <yadeyade .. fill in when the chapter is actually written.>
	% TODO: fill this part in!!!
	\item Every NI method has certain topological biases. We provide a tool for analysing the topological properties of large, evolving networks and use this to iteratively optimise GRN predictions (Chapter~\ref{chap:incgraph}).
	\item We discuss reproducibility problems of TI methods due to low rates of quantitative self-assessment (Chapter~\ref{chap:selfassessment}). We provide solutions for different causal reasons for this phenomenon in order to spur developers to perform more self-assessments.
	\item Finally, we summarise our experience in benchmarking computational methods in a list of essential guidelines (Chapter~\ref{chap:guidelines}).
\end{itemize}

@@

\begin{itemize}
	\item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. Computational methods for trajectory inference from single-cell transcriptomics. European journal of immunology. 2016 Nov;46(11):2496-506.
	\item \textbf{Cannoodt R}, Ruyssinck J, Ramon J, De Preter K, Saeys Y. IncGraph: Incremental graphlet counting for topology optimisation. PloS one. 2018 Apr 26;13(4):e0195997.
	\item Saelens W *, \textbf{Cannoodt R} *, Todorov H, Saeys Y. A comparison of single-cell trajectory inference methods. Nature biotechnology. 2019 May;37(5):547.
	\item \textbf{Cannoodt R}, Saelens W, Sichien D, Tavernier S, Janssens S, Guilliams M, Lambrecht B, De Preter K, Saeys Y. SCORPIUS: Fast, accurate, and robust single-cell pseudotime. In preparation.
	\item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyngen: Simulating developing single cells. In preparation.
	\item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyno: A toolkit for inferring, visualising, and interpreting trajectories. In preparation.
	\item \textbf{Cannoodt R}, Saelens W, Saeys Y, De Preter K. bred: Inferring single cell regulatory networks. In preparation. \fbox{\color{red}order authors?}
	\item \textbf{Cannoodt R}, Saelens W, Saeys Y. Self-assessment in trajectory inference. In preparation.
\end{itemize}

@@

Packages that were created as part of this work are hosted on Github under the username rcannood\footnote{\url{https://github.com/rcannood?tab=repositories}} or the dynverse organisation\footnote{\url{https://github.com/dynverse?tab=repositories}}. As part of our standard development practices, we automate execution of unit tests and writing extensive documentation to ensure the code complies with CRAN policy before submission. We aim to submit all other packages to CRAN as well.

>>>>

During the process of \textbf{transcription} that takes place in the cell nucleus, a complementary RNA copy is transcribed from the template DNA. The initial RNA transcript is a precursor messenger RNA (pre-mRNA) that needs to undergo series of maturation steps to ultimately form the mature messenger RNA (mRNA). This maturation includes pre-mRNA \textbf{splicing} to remove non-protein coding intervening sequences (the introns) and to join the neighbouring protein-coding sequences (the exons). A single pre-mRNA can be alternatively spliced to generate multiple forms of mRNAs that will result in the production of multiple protein isoforms. This process of alternative splicing is essential to generate more than 100'000 different proteins starting from just 20'000 genes\cite{nilsen_expansioneukaryoticproteome_2010}.

@@

\subsection{Cell types}
Ever since Robert Hook first described the different structures of cells in 1665, biologists have been classifying cells by form and function.
The human body is said to contain more than 210 different cell types that are classified into four groups: epithelial, connective, muscle, and nervous. This however, is a major underestimation of the real number of cell types. Neurons, for instance, that are known to be extremely diverse, are estimated to reach numbers above 10,000 different types\cite{strachan_humanmoleculargenetics_2011}.

@@

Flow cytometry\cite{fulwyler_electronicseparationbiological_1965} circumvents imaging and segmentation issues by measuring fluorescently labelled proteins as cells pass through a fluidic system.
Since cells need to be suspended in a buffer, flow cytometry is particularly useful for analysing non-adherent cells such as the many different immune cells in blood. However, many protocols already exist to extract viable single cells from tissues and tumours\cite{leelatian_preparingviablesingle_2017}.
Conventional flow cytometry devices enable to measure protein expression levels of millions of cells using up to eight different antibody fluorochromes simultaneously, while state-of-the-art instrumentation allows detection of up to 27 biomarkers simultaneously\cite{cossarizza_guidelinesuseflow_2017}.

@@

\subsubsection{Omics}
On the other side of the spectrum are the so-called "omics" technologies. "Omics"\footnote{The etymology of "omics" is quite interesting\cite{yadav_wholenesssuffixomics_2007}.} is a collective term for profiling all molecules of a particular type in a high-throughput manner.
There are at least ten types of "omics". In this work, we mostly consider genomics, transcriptomics, proteomics, and regulomics. Genomics studies the complete DNA sequence of an organism's genome, while transcriptomics and proteomics study the RNA transcripts and proteins, respectively. Regulomics studies the regulatory molecules (e.g. genes, RNAs, proteins) which play a role in determining gene regulation.

@@

The rapidly advancing field of single-cell omics harbours exceptional opportunities to discover new aspects of biology and redefine existing knowledge. Some of these opportunities lie in efforts like the HCA consortium\cite{regev_humancellatlas_2018}. They have set out to redefine all human cell types in both terms of their gene expression and location, and as well as the developmental trajectories connecting the different cell types. As part of this endeavour, the consortium will likely profile the whole transcriptomes of tens or even hundreds of millions of cells\cite{humancellatlasconsortium_humancellatlas_2018}.

@@

\subsection{Trajectory inference}
While clustering methods divide cells into distinct groups, trajectory inference (TI) methods acknowledge that cells are dynamic entities which transition from once cellular state to another via various dynamic processes. Rather than making distinct groups, TI methods allow studying dynamic processes by reconstructing the topology of a dynamic process as a trajectory, and map the cells onto that trajectory. A trajectory is a graph where the nodes represent noteworthy cellular states, and each cell is predicted to be progressing along transitions between the different states (Figure~\ref{fig:comp_tools_ti}A).

A trajectory can be visualised as a graph to better highlight the topology of the trajectory(Figure~\ref{fig:comp_tools_ti}A middle), as a heatmap to better depict the changes in gene expression along the different transitions (Figure~\ref{fig:comp_tools_ti}A right), or may be embedded in a dimensionality reduction of the cells to better demonstrate cellular heterogeneity along the trajectory (Figure~\ref{fig:comp_tools_ti}B right). Similar to clustering, by colouring the cells according to the expression of genes known to be involved in the dynamic process of interest, the milestones in the trajectory can be annotated (Figure~\ref{fig:comp_tools_cluster}B).

\begin{figure}[htb!]
	\centering
	\includegraphics[width=.9\linewidth]{fig/comptools2/comptool_3_trajectoryinference.pdf} 
	\caption{
		\textbf{Trajectory inference for single-cell omics data.}
		\textbf{A:} During a dynamic process cells pass through several transitional states, characterized by different waves of transcriptional, morphological, epigenomic and/or surface marker changes\cite{enver_stemcellstates_2009}. TI methods provide an unbiased approach to identifying and correctly ordering different transitional stages. 
		\textbf{B:} By overlaying gene expression levels on a dimensionality reduction, the milestones can be annotated to allow better interpretation of the cellular heterogeneity.
	}
	\label{fig:comp_tools_ti}
\end{figure}

A lot of TI methods use similar algorithms to be able to infer a trajectory. By breaking down each method into its set of core algorithms, we can create a map of TI methodology\cite{cannoodt_computationalmethodstrajectory_2016} (Figure~\ref{fig:ti_metro}A), which is divided into two major classes.
In the first step, dimensionality reduction techniques such as manifold learning, clustering, or graph-based methods are used to convert the dataset to a more simplified representation. This representation of the data then allows the trajectory itself to be more easily modelled in a second step. In this second step, the trajectory is modelled within the data using graph-based or curve-based approaches, after which the cells themselves can be ordered using a variety of methods.

A common way to classify TI methods is by the types of trajectories they can infer\cite{saelens_comparisonsinglecelltrajectory_2019} (Figure~\ref{fig:ti_metro}B). About half of TI methods specialise in inferring linear or cyclic trajectories (i.e. they order the cells). Others model the trajectory as a rooted tree, allowing for one or more bifurcations to occur. Only a few methods are able to infer more generalised trajectories containing disconnected subgraphs or cycles.

\begin{figure}[htb!]
	\centering
	\includegraphics[width=.9\linewidth]{fig/ti_metro} 
	\caption{
		\textbf{TI methods use several common building blocks and can be organized in a unifying modular framework.} 
		\textbf{A:} Most TI methods consist of two major steps, dimensionality reduction and trajectory modelling. TI methods   require some form of dimensionality reduction in order to summarise cell heterogeneity in a lower dimensional space. Subsequently, a trajectory modelling step then operates in this reduced space, aiming to identify cell states, constructing a trajectory through the different states, and projecting the cells back on to the trajectory. 
		\textbf{B:} TI methods can be classified according to the trajectory topologies they can infer.
	}
	\label{fig:ti_metro}
\end{figure}

@@

Network inference (NI) methods predict which genes are regulated by which transcription factors (Figure~\ref{fig:comp_tools_ni}). The output of a network inference is thus a graph, where nodes represent genes and edges denote a regulatory interaction between a regulator and a target gene. Interactions typically have two properties: its regulatory strength (a positive real value) and its effect (promoting or repressing).

Before single-cell omics, these methods rely on multiple experiments, amongst which perturbation and time-series experiments, to predict the effect each transcription factor has on the up- or downregulation of a gene. One of the main advantages of single-cell omics is the heterogeneity between cells caused by naturally occurring biological randomness\cite{padovan-merhar_usingvariabilitygene_2013} can be exploited to infer regulatory interactions between TFs and their target genes at much lower costs.

@@

\section{Research context and objectives} \label{sec:research_objectives}
Recent technological advancements in profiling single cells are having significant repercussions in many fields of biology. Profiling thousands of individual cells in a genome-wide manner provides opportunities to study cell heterogeneity and dynamics, for example inferring mechanisms for cellular development or intercellular communication. 
Hundreds of new software tools were developed\cite{zappia_exploringsinglecellrnaseq_2018} to perform these new types of analyses, or to fit existing analytical tools to deal with new data characteristics (e.g. differential expression, dimensionality reduction, normalisation).

One major shortcoming during the advent of single-cell omics was that the majority of newly developed computational tools were not quantitatively and comparatively evaluated. Rather, they relied on anecdotal evidence to demonstrate its usefulness. This issue is not the result of the tool developer's malevolence, but instead of the lack of data required to perform such comprehensive benchmarks.
%TODO: also mention guidelines for users.

@@

In this work, we aim to speed up scientific progress in single-cell omics by providing tools both for end-users and developers alike. For developers of computational approaches, we provide tools and guidelines for benchmarking their method on real and synthetic data. For end-users we develop new tools and guidelines for analysing dynamic processes by inferring trajectories and gene regulatory networks. Our specific scientific objectives are discussed in the following chapters.

\begin{itemize}
	\item We \textbf{develop benchmarking strategies} for assessing the performance of computational tools constrained by low availability of novel types of real single-cell data (Chapter~\ref{chap:dyngen}). \textit{In silico} simulations of individual cells are used to help kick-start emerging domains much more safely and allow anticipation of future technological developments by already developing computational tools.
	\item We apply this strategy to \textbf{perform a comparison of TI methods} (Chapter~\ref{chap:dynbenchmark}). Trajectory inference is one of the largest categories of all the novel single-cell omics tools, yet a comprehensive and quantitative study of the advantages and disadvantages of the numerous tools was hitherto lacking. We provide a set of guidelines for end-users wishing to infer trajectories. Our pipeline, datasets, metrics, and containerised wrappers of TI methods are made publicly available for developers to use.
	\item We develop \textbf{a toolkit to infer, visualise and interpret single-cell trajectories} using more than 50 different TI methods (Chapter~\ref{chap:dyno}). dyno provides downstream analysis such as: visualising a trajectory in a low-dimensional space or a heatmap, detecting genes differentially expressed at different stages of the tragectory, comparing multiple trajectories in a common dimensionality reduction, and manipulating the trajectory (e.g. adding directionality or adding annotation).
	\item We introduce a TI method specialised in \textbf{inferring linear trajectories} (Chapter~\ref{chap:scorpius}). Despite linear TI being the most simple but commonly used form of trajectory inference, the benchmark demonstrated that most TI methods are not capable of producing accurate models of linear datasets.
	\item We develop an algorithm for  \textbf{inferring gene regulatory interactions at a single-cell level} (Chapter~\ref{chap:bred}). We demonstrate our NI methodology by inferring regulatory interactions for 25'000 hematopoietic cells, as well as for 15'000 cancer-related bulk profiles. % TODO: was this chapter written in the end?
	\item We provide a tool for  \textbf{analysing the topological properties of large, evolving networks} and use this to iteratively optimise GRN predictions (Chapter~\ref{chap:incgraph}).
	\item We reflect on \textbf{challenges in performing self-assessments of TI methods} (Chapter~\ref{chap:selfassessment}).
	Reproducability problems of TI methods are partially due to low rates of quantitative self-assessment. We look at 75 manuscripts of TI methods and identify key challenges the authors had in benchmarking their tool. For each of these challenges, we provide a summary of how we and others have successfully tackled these challenges, in order to spur developers to perform more self-assessments.
	\item Finally, we \textbf{summarise our experience in benchmarking computational methods} by providing a list of essential guidelines in benchmarking computational tools (Chapter~\ref{chap:guidelines}).
\end{itemize}

@@

\begin{itemize}
  \item \textbf{Cannoodt R}, Saelens W, Sichien D, Tavernier S, Janssens S, Guilliams M, Lambrecht B, De Preter K, Saeys Y. SCORPIUS improves trajectory inference and identifies novel modules in dendritic cell development. bioRxiv 079509. 2016 Oct.
	\item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. Computational methods for trajectory inference from single-cell transcriptomics. European journal of immunology. 2016 Nov;46(11):2496-506.
	\item \textbf{Cannoodt R}, Ruyssinck J, Ramon J, De Preter K, Saeys Y. IncGraph: Incremental graphlet counting for topology optimisation. PloS one. 2018 Apr 26;13(4):e0195997.
	\item Saelens W *, \textbf{Cannoodt R} *, Todorov H, Saeys Y. A comparison of single-cell trajectory inference methods. Nature biotechnology. 2019 May;37(5):547.
	\item \textbf{Cannoodt R}, Saelens W, Sichien D, Tavernier S, Janssens S, Guilliams M, Lambrecht B, De Preter K, Saeys Y. SCORPIUS: Fast, accurate, and robust single-cell pseudotime. In preparation.
	\item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyngen: Simulating developing single cells. In preparation.
	\item \textbf{Cannoodt R} *, Saelens W *, Saeys Y. dyno: A toolkit for inferring, visualising, and interpreting trajectories. In preparation.
	\item \textbf{Cannoodt R}, Saelens W, Saeys Y, De Preter K. bred: Inferring single cell regulatory networks. In preparation. 
	\item \textbf{Cannoodt R}, Saelens W, Saeys Y. Self-assessment in trajectory inference. In preparation.
\end{itemize}

@@

Packages that were created as part of this work are hosted on Github under the username rcannood\footnote{\url{https://github.com/rcannood?tab=repositories}} or the dynverse organisation\footnote{\url{https://github.com/dynverse?tab=repositories}}. As part of our standard development practices, we automate execution of unit tests and write extensive documentation to ensure the code complies with CRAN policy before submission. We aim to submit all other packages to CRAN as well.

### 2019-10-14 ch6_bred/content.tex
update

>>>>

One of the central cellular processes underlying development is transcriptional regulation. During development, changes in transcription factor activity induce chromatin modifications, chromatin remodeling and ultimately a differential recruitment of the basal transcriptional machinery [66]. Modeling the dynamics of gene regulation is therefore essential to better understand why a cellular dynamic processes progresses through several steps, and what goes wrong in the case of disease.

The dynamics of gene regulation has classically been studied using time series data [23]. When dynamic processes progress asynchronously, such as in hematopoiesis, time series data are usually obtained by sorting different transition states and assessing bulk gene expression and transcription factor binding within the population [67–70]. Alternatively, time series data can also be generated by synchronizing the dynamic process between cells. However, issues with time-resolution, heterogeneity and good in vivo synchronization models can often limit the predictive power of the dynamic models of gene regulation which can be constructed [23]. Combining single-cell snapshot data with trajectory inference can now in principle construct an accurate time-series dataset for every cell population both in vivo and in vitro, while avoiding issues with heterogeneity. Initially, several pioneering studies demonstrated the potential of this approach in single cells using single-cell qPCR data for selected sets of 10 to 50 transcription factors [71–75]. With the advent of genome wide single-cell technologies, the models of dynamic gene regulation can now be made even more complete while avoiding a prior bias to well studied transcription factors [17,30,76,77] and several methods have already been developed specifically tuned towards single-cell data [78–80].

The power of a dynamic network approach in single cells is not only that it allows the identification of transcription factors specifically regulating a specific state, but also how these factors then induce sets of genes necessary for regulating the next wave of transcription changes (Figure 1B). Such a gene regulatory cascade was described during cellular reprogramming from mouse embryonic fibroblasts to induced pluripotent stem cells, where SOX2 was shown to activate a series of other transcription factors which, together with SOX2, induces pluripotency [72]. Transition states are frequently associated with particular subnetworks of transcriptional regulators, as was shown in a direct reprogramming setting from fibroblasts to neurons [77]. Several studies have highlighted how some regulatory links can be very dynamic while others show evidence of being static during consecutive developmental stages [74,75]. Regulatory network motifs can also be extracted from the regulatory network [74], some of which can have interesting properties in dynamic systems [81,82]. In branching trajectories dynamic regulatory networks can also be used to study the regulators important for a particular branch [17,75,79] (Figure 1C).

\begin{figure}[htb!]
	\centering
	\includegraphics[width=\linewidth]{fig/tcga/grouped_interactions.pdf} 
	\caption{
		A
	}
	\label{fig:tcga}
\end{figure}

### 2019-10-14 ch8_discussion/content.tex
clean up thesis
update

\blindtext

\section{Overview and conclusions of the presented work}

% notice that normalisation was absent from this dissertation.
%Computer scientists should proceed with caution in developing new software, however, as the results produced should not only be convincing and easy to interpret, but the software should be robust and generate sufficiently accurate models of the underlying system.
%
%Bit too excited -- false positives, poor accuracy, scalability issues, poor software quality.

\section{Future research directions}

>>>>

% DISCUSSION CONTENT

In this work, we aimed to accelerate scientific progress in the field of single-cell omics by providing and evaluating computational tools. In section~\ref{sec:research_objectives} we proposed a concrete set of objectives this work will attempt to solve. Every chapter respectively tackles one of these research objectives, and each chapter discusses the implications of that part of our research in detail. This chapter reflects on the impact of this work on the field, going over each of the objectives and respective chapters.

\section{Benchmarking with \textit{in silico} single cells}

We developed \texttt{dyngen}, a simulator of single cells to benchmark and stress-test computational tools for single-cell omics (Chapter~\ref{chap:dyngen}).
It supports a selection of existing experimental protocols, but can also simulate experiments for permitted by current technologies, such as generating omics profiles from the same cell at multiple timepoints.
This data can help kick-start emerging domains with low data availability more safely by allowing method developers to test their method before numerous datasets become publicly available.

\texttt{dyngen} has already been used successfully to evaluate trajectory inference\cite{saelens_comparisonsinglecelltrajectory_2019}, trajectory alignment\cite{vandenberge_trajectorybaseddifferentialexpression_2019}, and network inference\cite{pratapa_benchmarkingalgorithmsgene_2019} methods. Furthermore, we showed that it can also be used to evaluate differential network inference methods, and trajectory alignment methods.

\section{Comparing 45 trajectory inference methods}
Using this simulator and a collection of real datasets, we performed a comparison of 45 TI methods (Chapter~\ref{chap:dynbenchmark}).
Our contributions include writing software to run 45 different error-prone TI methods with a common interface, downloading and processing hundreds of single cell datasets, and developing novel metrics for comparing ground truth and predicted trajectories.
With this benchmarking study, we had two clear goals in mind: to help guide users to TI methods that are suitable for their needs, and to provide developers of TI methods with the necessary tools to benchmark their own tool.

We accomplished the first goal by constructing a set of guidelines for end-users to help choose a TI method that is accurate, robust and fit for their application. 
Since such guidelines were hitherto lacking, they are now commonly disseminated in manuscripts \cite{lafzi_tutorialguidelinesexperimental_2018,luecken_currentbestpractices_2019}, courses \cite{kiselev_analysissinglecell_2019,martens_analysissinglecell_2019}, and slides shown during keynote caffeine refuelling sessions \cite{hemberg_coffeebreakanalysis_2019}.

For our second goal, it seems too early to tell whether our research had any effect on self-assessments of TI method developers. We did made our pipeline, datasets, metrics, and containerised wrappers of TI methods publicly available for developers to use.
However, so far we do not observe an increase in developers performing quantitative benchmarks. We hypothesise causal reasons for this phenomenon and provide solutions in order to spur TI developers to perform more self-assessments (Chapter~\ref{chap:selfassessment}).

\section{A toolkit to infer, visualise and interpret single-cell trajectories}

The previous two works necessitated developing a tool for visualising trajectory data. We extended this implementation into a full toolkit, named \texttt{dyno}, for inferring and analysing trajectories (Chapter~\ref{chap:dyno}). The toolkit allows performing downstream analyses such detecting differentially expressed genes along transitions in the trajectory, annotating the trajectory, and comparing multiple trajectories in a common dimensionality reduction. A major benefit of \texttt{dyno} is that it allows utilising any of the 50 TI methods we wrapped as part of the comparison, without having to install dependencies for every TI method separately.

The software packages consist of several subpackages, each with a separate functionality (e.g. \texttt{dynplot} visualises trajectories, \texttt{dynmethods} provides wrappers for the TI methods). Since most of the packages are currently only hosted on Github, only users familiar with Github have been able to use it. To make \texttt{dyno} available to a larger audience, we are publishing each of these packages and developing command-line interfaces for the most important functionalities.

\section{Fast, accurate, and robust single-cell pseudotime}

We developed a TI method, called \texttt{SCORPIUS}, specialised in inferring linear trajectories (Chapter~\ref{chap:scorpius}). Since our comparison of TI methods already lists 30 TI methods that can infer trajectories that are more complex than a linear ordering of cells, why would we need a linear TI method?

Being able to study how cells progress over time is an analysis as fundamental as being able to cluster cells or identify differentially expressed genes between two groups. We showed that for analysing datasets containing linear trajectories, \texttt{SCORPIUS} outperforms all other TI methods tested as part of our TI benchmark.

Despite \texttt{SCORPIUS} published on bioRxiv in 2016, discussion and usage of the pre-print has largely been limited to comparing the results of a different (linear) TI methods to those of \texttt{SCORPIUS}. Since the pre-print, many improvements have been made to improve the accuracy and robustness of predicted trajectories, to allow scaling up to datasets containing 100'000s of cells, and to providing a more user-friendly interface. Since the beginning of 2019 -- before the publication of our benchmark of TI methods -- several studies have now reported using SCORPIUS as a tool for deriving primary results\cite{damond_maphumantype_2019, cheng_multifactorialheterogeneityvirusspecific_2019, tibbitt_singlecellrnasequencing_2019, hove_singlecellatlasmouse_2019, wouters_singlecellgeneregulatory_2019}.

\section{bred: Inferring single cell regulatory networks}

%	\item We develop an algorithm for  \textbf{inferring gene regulatory interactions at a single-cell level} (Chapter~\ref{chap:bred}). We demonstrate our NI methodology by inferring regulatory interactions for 25'000 hematopoietic cells, as well as for 15'000 cancer-related bulk profiles. % TODO: was this chapter written in the end?

%\item We invent a new type of NI method capable of inferring the GRN of individual cells (Chapter~\ref{chap:bred}). We demonstrate this <yadeyade .. fill in when the chapter is actually written.>

\section{incgraph: Optimising regulatory networks}

%	\item We provide a tool for  \textbf{analysing the topological properties of large, evolving networks} and use this to iteratively optimise GRN predictions (Chapter~\ref{chap:incgraph}).

%\item Every NI method has certain topological biases. We provide a tool for analysing the topological properties of large, evolving networks and use this to iteratively optimise GRN predictions (Chapter~\ref{chap:incgraph}).

\section{A life without Git, Travis CI, or tidyverse}
A significant portion of this work involved developing large software libraries, in a collaborative setting, over a time span of about four years. Since the results of our comparison of 45 TI methods required three years to develop, we were happily forced to develop a system where experiments could be rerun and results could be updated on-the-fly.

We summarised our experiences in the form of a set of guidelines for benchmarking computational tools (Chapter~\ref{chap:guidelines}). However, these guidelines do not touch upon many aspects of good software development practices that we learned to use in order to bring this thesis to a good end. In particular, I would like to highlight several  fundamental (open-source) projects without the likes of which our research would have been simply impossible to perform, namely Git, Travis CI, and the tidyverse.

\textbf{Git}\cite{torvalds_gitfastversion_2005} is a code-revision system that allows multiple users to collaborate on developing code and keep track of what changes were made by whom. 
Since Wouter and I were often working closely together on a specific part of our software, Git saved us a lot of time by merging together changes developed in parallel. Only in few cases did we need to intervene manually to merge our changes.
Additionally, Github.com allowed us not only to collaborate with each-other, but also correspond and collaborate with other software developers from many different research groups, using Github Issues to discuss problems with other researchers, and Github Pull Requests to contribute code to open-source projects.

Together, we published over 20'000 code contributions (commits) across 50+ software packages\cite{cannoodt_developmentdynverse_2019}. Since many of these packages depend on one another, it is inevitable that changes made to one package would break another package. We wrote many, many unit tests (and still too few), and let these tests automatically be executed on \textbf{Travis CI}\cite{traviscigmbh_traviscitest_2011}. This way, when we pushed breaking changes to Github, Travis CI would notify that our commit has resulted in one or more unit tests failing. While sometimes it is a hassle to set up, Travis CI has prevented us countless times from generating faulty results due to a faulty underlying function.

Software bugs can be introduced even by incredibly small, seemingly insignificant changes. The R programming language seems particularly susceptible towards getting trapped by its many pitfalls. However, there are many benefits of using R in bioinformatics research context, such as its extensive community of statisticians develop packages for the R ecosystem. The \textbf{tidyverse} packages\cite{wickham_tidytoolsmanifesto_2017}, developed by the RStudio team and many online contributors, completely transformed my experiences with R from tedious struggles to efficient everyday functional programming.

Honourable mentions go to Linux, Fedora, \LaTeX, TeXstudio, (R)Markdown, Bash, Sed, Regular expressions, Rocket Chat, for making bioinformatics software development even more fun and enjoyable.

\clearpage
\section{References}
\printbibliography[heading=none]
